---
title: "Data in Motion: Data Analysis Challenge Week 4"
output:
  html_document:
    df_print: paged
  html_notebook: default
  pdf_document: default
  word_document: default
---
```{r}
# loading the packages
library(tidyverse)
library(readr)
library(ggplot2)
library(dplyr)
library(lubridate)
```

```{r}
# importing the dataset
spotify <- read.csv("Most Streamed Spotify Songs 2024.csv")
```

```{r}
head(spotify)
```

#### Data Cleaning
```{r}
# function to clean comma and convert to numeric
clean_numeric <- function(x) {
  x <- gsub(",", "", x)  # Remove commas
  as.numeric(x)  # Convert to numeric
}
# apply the function to pandora streams
spotify$Pandora.Streams <- clean_numeric(spotify$Pandora.Streams)
class(spotify$Pandora.Streams)

# apply the function to sound cloud streamns
spotify$Soundcloud.Streams <- clean_numeric(spotify$Soundcloud.Streams)
class(spotify$Soundcloud.Streams)

# apply function to spotify playlist count
spotify$Spotify.Playlist.Count <- clean_numeric(spotify$Spotify.Playlist.Count)
class(spotify$Spotify.Playlist.Count)

# apply function to spotify playlist reach
spotify$Spotify.Playlist.Reach <- clean_numeric(spotify$Spotify.Playlist.Reach)
class(spotify$Spotify.Playlist.Reach)

# apply function to shazam count
spotify$Shazam.Counts<- clean_numeric(spotify$Shazam.Counts)
class(spotify$Shazam.Counts)

# apply function to siriusxm
spotify$SiriusXM.Spins<- clean_numeric(spotify$SiriusXM.Spins)
class(spotify$SiriusXM.Spins)

# convert explicit tract to facror
spotify$Explicit.Track<- as.factor(spotify$Explicit.Track)
class(spotify$Explicit.Track)

```
```{r}
class(spotify$Release.Date)
```
```{r}
# convert release date to date type
spotify$Release.Date <- mdy(spotify$Release.Date)
```
```{r}
class(spotify$Release.Date)
```
```{r}
# extracting month, year from the release date 
spotify <- spotify %>% 
  mutate(Release.Month = month(Release.Date,label = TRUE, abbr = TRUE)) %>%
  mutate(Release.Year = year(Release.Date))
```


#### Trend Analysis

##### Identify and analyze the trends in song releases. Are there specific months or seasons when most top-streamed songs are released? What patterns can be observed?
```{r}
# songs released in each year
released_songs_year <- spotify %>% 
  group_by(Release.Year) %>% 
  summarise(total_songs = n()) %>% 
  arrange(desc(total_songs))
tibble(released_songs_year)

# albums release by year
album_year <- spotify %>% 
  group_by(Release.Year) %>% 
  summarise(total_albums = n_distinct(Album.Name)) %>% 
  arrange(desc(total_albums))
tibble(album_year)
```
```{r}
# top 10 spotify streamed songs and the release year,month
top_10_spotify_streams <- spotify %>% 
  group_by(Track) %>% 
  summarise(stream = sum(Spotify.Streams),
            Year = first(Release.Year),
            Month = first(Release.Month)) %>% 
  mutate(streams = stream/1e9,  # convert to billions
         streams = sprintf("%.2fB",streams)) %>% # format as strings
  arrange(desc(stream)) %>% 
  select(Track,streams,Month,Year)

top_10_spotify_streams <-tibble(head(top_10_spotify_streams, 10))
top_10_spotify_streams
```
```{r}
# top 10 pandora streamed songs and the release year,month
top_10_pandora_streams <- spotify %>% 
  group_by(Track) %>% 
  summarise(stream = sum(Pandora.Streams),
            Year = first(Release.Year),
            Month = first(Release.Month)) %>% 
   mutate(streams = stream/1e9,  # convert to billions
         streams = sprintf("%.2fB",streams)) %>% # format as strings
  arrange(desc(stream)) %>% 
  select(Track,streams,Month,Year)

 top_10_pandora_streams <-tibble(head(top_10_pandora_streams, 10))
 top_10_pandora_streams
```
```{r}
# top 10 soundcloud streamed songs and the release year,month
top_10_soundcloud_streams <- spotify %>% 
  group_by(Track) %>% 
  summarise(stream = sum(Soundcloud.Streams),
            Year = first(Release.Year),
            Month = first(Release.Month)) %>% 
   mutate(streams = stream/1e6,  # convert to millions
         streams = sprintf("%.2fM",streams)) %>% # format as strings
  arrange(desc(stream)) %>% 
  select(Track,streams,Month,Year)

top_10_soundcloud_streams <-tibble(head(top_10_soundcloud_streams, 10))
top_10_soundcloud_streams
```
```{r}
# overall top streamed songs on all platforms
top_10_streamed_songs <- spotify %>% 
  mutate(total_streams = rowSums(select(.,Spotify.Streams,Pandora.Streams,Soundcloud.Streams))) %>% #adding streams from all platforms 
  group_by(Track) %>% 
  summarise(stream = sum(total_streams, na.rm = TRUE),
           Month = first(Release.Month) ) %>% 
  mutate(streams = stream/1e9,  # convert to billions
         streams = sprintf("%.2fB",streams)) %>% # format as strings
  arrange(desc(stream)) %>% 
  select(Track,streams,Month)
top_10_streamed_songs <-tibble(head(top_10_streamed_songs, 10))

top_10_streamed_songs 
```

#### Artist and Genre Impact

##### Determine the impact of artists and genres on the streaming counts. Which artists have the highest average streams?

```{r}
# artists with the highest average streams on spotify
top_10_spotify_artists <- spotify %>% 
  group_by(Artist) %>% 
  summarise(avg_streams = mean(Spotify.Streams)) %>% 
   mutate(streams = avg_streams/1e9,  # convert to billions
         streams = sprintf("%.2fB",streams)) %>% # format as strings
  arrange(desc(avg_streams)) %>% 
  slice_head(n = 10) %>% 
  select(Artist,streams)
top_10_spotify_artists

# artists with the highest average streams on pandora
top_10_pandora_artists <- spotify %>% 
  group_by(Artist) %>% 
  summarise(avg_streams = mean(Pandora.Streams)) %>% 
  mutate(streams = avg_streams/1e9,  # convert to billions
         streams = sprintf("%.2fB",streams)) %>% # format as strings
  arrange(desc(avg_streams)) %>% 
  slice_head(n = 10) %>% 
 select(Artist,streams)
top_10_pandora_artists

# artists with the highest average streams on soundcloud
top_10_soundcloud_artists <- spotify %>% 
  group_by(Artist) %>% 
  summarise(avg_streams = mean(Soundcloud.Streams)) %>% 
  mutate(streams = avg_streams/1e6, # convert to millions
      streams = sprintf("%.2fM",streams)) %>% # format as strings
  arrange(desc(avg_streams)) %>% 
  slice_head(n = 10) %>% 
  select(Artist,streams)
top_10_soundcloud_artists

# artists with the highest average streams on all platforms
top_10_overall_artist <- spotify %>% 
  rowwise() %>%      # calculate the mean streams per song
  mutate(mean_streams = mean(c_across(c(Spotify.Streams,Pandora.Streams,Soundcloud.Streams)))) %>% 
  group_by(Artist) %>%  # group by artists and calculate avg mean streams
  summarise(streams = mean(mean_streams, na.rm = TRUE)) %>% 
  mutate(streams_B = streams/1e9,  # convert to billions
         streams_B = sprintf("%.2fB",streams_B)) %>% # format as strings
  arrange(desc(streams)) %>% 
  slice_head(n = 10) %>% 
  select(Artist, streams_B)
top_10_overall_artist
```
#### Playlist Influence
##### Assess the influence of playlist inclusion on the streaming counts. How does being featured in playlists correlate with the total streams of a song? 
```{r}
# calculate the correlation between spotify streams and the playlist count
correlation <- cor(spotify$Spotify.Streams, spotify$Spotify.Playlist.Count, use = "complete.obs")

# Print correlation coefficients
print(paste("Correlation between Spotify Streams and Spotify Playlist Count: ", correlation))
```
#### External Factors
##### Investigate how external factors such as explicit content, Shazam counts, and SiriusXM spins affect the popularity of songs. Is there a significant correlation between these factors and the streaming counts?

```{r}
# run a multiple regresssion analysis to examine the examine the relationship between Spotify streams and explicit track,Shazam counts, and SiriusXM spins.

model <- lm(Spotify.Streams ~Explicit.Track + Shazam.Counts
            + SiriusXM.Spins,data = spotify)
summary(model)
```
```{r}
# Calculate correlation coefficients
cor_shazam <- cor(spotify$Spotify.Streams, spotify$Shazam.Counts, use = "complete.obs")
cor_siriusxm <- cor(spotify$Spotify.Streams, spotify$SiriusXM.Spins, use = "complete.obs")

# Print correlation coefficients
print(paste("Correlation between Spotify Streams and Shazam Counts: ", cor_shazam))
print(paste("Correlation between Spotify Streams and SiriusXM Spins: ", cor_siriusxm))

```







